---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
```

# VarReg

<!-- badges: start -->
`r badge_cran_release("VarReg", "orange")` 
`r badge_devel("kristyrobledo/VarReg", "blue")`
`r badge_cran_download("VarReg", "grand-total", "blue")`
`r badge_cran_download("VarReg", "last-month", "green")`
`r badge_cran_download("VarReg", "last-week", "yellow")`
<!-- badges: end -->

<!-- Please don't remove this: Grab your social icons from https://github.com/carlsednaoui/gitsocial -->

<!-- display the social media buttons in your README -->

<!--[![alt text][1.1]][1] -->



<!-- links to social media icons -->
<!-- no need to change these -->

<!-- icons with padding -->

[1.1]: http://i.imgur.com/tXSoThF.png (twitter icon with padding)
[2.1]: http://i.imgur.com/P3YfQoD.png (facebook icon with padding)
[3.1]: http://i.imgur.com/yCsTjba.png (google plus icon with padding)
[4.1]: http://i.imgur.com/YckIOms.png (tumblr icon with padding)
[5.1]: http://i.imgur.com/1AGmwO3.png (dribbble icon with padding)
[6.1]: http://i.imgur.com/0o48UoR.png (github icon with padding)

<!-- icons without padding -->

[1.2]: http://i.imgur.com/wWzX9uB.png (twitter icon without padding)
[2.2]: http://i.imgur.com/fep1WsG.png (facebook icon without padding)
[3.2]: http://i.imgur.com/VlgBKQ9.png (google plus icon without padding)
[4.2]: http://i.imgur.com/jDRp47c.png (tumblr icon without padding)
[5.2]: http://i.imgur.com/Vvy3Kru.png (dribbble icon without padding)
[6.2]: http://i.imgur.com/9I6NRUm.png (github icon without padding)


<!-- links to your social media accounts -->
<!-- update these accordingly -->


<!-- Please don't remove this: Grab your social icons from https://github.com/carlsednaoui/gitsocial -->

The goal of VarReg is to provide methods for fitting semi-parametric mean and variance models, with normal or censored data. This has also been extended to allow a regression in the location, scale and shape parameters.
This algorithm is based upon an EM (Expectation Maximisation) algorithm, so is more stable than other similar methods like GAMLSS.

## :raising_hand: Author

Kristy Robledo <https://github.com/kristyrobledo>

NHMRC Clinical Trials Centre, University of Sydney


## :arrow_double_down: Installation

You can install the released version of VarReg from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("VarReg")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kristyrobledo/VarReg")
```
## :book: Examples

This is a basic example to read in the mcycle dataset and perform a linear model in the mean and the variance:

```{r example}
library(VarReg)
## read in dataset
data(mcycle)
## run a model with linear mean and linear variance:
linmodel<-semiVarReg(mcycle$accel, mcycle$times, meanmodel="linear", varmodel="linear",
  maxit=10000)

```

Now we can plot the model:

```{r plotlinear}
plotVarReg(linmodel)


##can also add CI
plotVarReg(linmodel, ci=TRUE, ci.type = "im")
```

Or we can look at the results:
```{r seelin}
linmodel$loglik

linmodel$mean

linmodel$variance
```

We can also run a model with semi-parametric mean (4 internal knots) and semi-parametric variance (2 knots):

```{r semimodel}
semimodel<-semiVarReg(mcycle$accel, mcycle$times, meanmodel="semi", varmodel="semi",
knots.m=4, knots.v=2, maxit=10000)
plotVarReg(semimodel)

## run a model with semi-parametric mean (4 internal knots) and semi-parametric monotonic
## variance (2 knots):
## not run
##semimodel_inc<-semiVarReg(mcycle$accel, mcycle$times, meanmodel="semi", varmodel="semi",
##knots.m=4, knots.v=2, mono.var="inc")
```

Lastly, we can fit a model with a model in the location, scale and shape. Im not going to run this, just show the code, as it takes a while to run on my laptop!

```{r lss}
 ##  LSS model followed by the basic plot command
#lssmodel<-lssVarReg(mcycle$accel, mcycle$times,  locationmodel="linear", scale2model="linear", shapemodel="constant", maxit=10000)
#plotlssVarReg(lssmodel, xlab="Time in seconds", ylab="Acceleration")
```

Enjoy!

