---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# VarReg

<!-- badges: start -->
<!-- badges: end -->

The goal of VarReg is to provide methods for fitting semi-parametric mean and variance models, with normal or censored data. This has also been extended to allow a regression in the location, scale and shape parameters.
This algorithm is based upon an EM (Expectation Maximisation) algorithm, so is more stable than other similar methods like GAMLSS.

## Installation

You can install the released version of VarReg from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("VarReg")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kristyrobledo/VarReg")
```
## Example

This is a basic example to read in the mcycle dataset and perform a linear model in the mean and the variance:

```{r example}
library(VarReg)
## read in dataset
data(mcycle)
## run a model with linear mean and linear variance:
linmodel<-semiVarReg(mcycle$accel, mcycle$times, meanmodel="linear", varmodel="linear",
  maxit=10000)

```

Now we can plot the model:

```{r plotlinear}
plotVarReg(linmodel)


##can also add CI
plotVarReg(linmodel, ci=TRUE, ci.type = "im")
```

Or we can look at the results:
```{r seelin}
linmodel$loglik

linmodel$mean

linmodel$variance
```

We can also run a model with semi-parametric mean (4 internal knots) and semi-parametric variance (2 knots):

```{r semimodel}
semimodel<-semiVarReg(mcycle$accel, mcycle$times, meanmodel="semi", varmodel="semi",
knots.m=4, knots.v=2, maxit=10000)
plotVarReg(semimodel)

## run a model with semi-parametric mean (4 internal knots) and semi-parametric monotonic
## variance (2 knots):
## not run
##semimodel_inc<-semiVarReg(mcycle$accel, mcycle$times, meanmodel="semi", varmodel="semi",
##knots.m=4, knots.v=2, mono.var="inc")
```

Lastly, we can fit a model with a model in the location, scale and shape. Im not going to run this, just show the code as it takes a while to run on my laptop!

```{r lss}
 ##  LSS model followed by the basic plot command
#lssmodel<-lssVarReg(mcycle$accel, mcycle$times,  locationmodel="linear", scale2model="linear", shapemodel="constant", maxit=10000)
#plotlssVarReg(lssmodel, xlab="Time in seconds", ylab="Acceleration")
```

Enjoy!

