---
title: "VarReg: Basic usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VarReg: Basic usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo=TRUE,
  collapse = TRUE,
  comment = "#>"
)
```


This repository is for the data and code presented in the paper published here (to be added). 

## Install packages

Firstly, you will need to install the `VarReg` package and a few others to perform the code below. 

```{r cars, message=FALSE}
library(VarReg)
library(gamlss) #for the CD4 dataset
library(tidyverse)
library(kableExtra)
rna<-read.csv(file="rna.csv") #load rna dataset
```

## CD4 dataset

This dataset is located within the `gamlss` package. CD4 is a type of white blood cell, and in this dataset, it has been measured in uninfected children born from HIV-1 infected women. The dataset contains 609 measurements of CD4 cell counts
and the child's age at which the measurements were taken.

```{r}

data("CD4")

str(CD4)
```

Lets reproduce the graphic in the paper, showing that at younger ages there is more variation in the CD4 counts than at older ages, demonstrating heteroscedasticity.

```{r}
ggplot(data=CD4, aes(y=cd4, x=age)) +
  geom_point()+
  xlab("Age of child in years")+
  ylab("CD4 cell count")+
  theme_minimal()
```

### Linear model in mean and variance

Let us fit a linear model in the mean and variance model, like so


```{r}
cd4.linear<-semiVarReg(y = CD4$cd4, 
                       x=CD4$age, 
                       meanmodel = "linear", 
                       varmodel = "linear", 
                       maxit=10000)
```

key outputs for model fit:

```{r}
cd4.linear$aic
cd4.linear$bic
```

Key estimates from model:

```{r}
cd4.linear$mean
cd4.linear$variance
```

### Visualise the model with `plotVarReg` function:

```{r, fig.dim=c(6,6)}
plotVarReg(cd4.linear, xlab = "Age in years", ylab = "CD4 cell count")
```

### Searching for the optimal spline

Searching to a max of 7 knots in the mean and variance, with maximum iterations (`maxit`) of 100 to minimise time taken for the search:

```{r, echo=FALSE}
if (file.exists("results.rds")) {
  cd4.best <- readRDS("results.rds")

} else {

  # create the results dataset -> now using the saved code as time consuming
  cd4.best <- searchVarReg(y=CD4$cd4, 
                           x=CD4$age, 
                           maxknots.m = 7, 
                           maxknots.v = 7, 
                           selection="AIC", 
                           maxit=100)  #time-consuming function so not 1000 
  saveRDS(cd4.best, "results.rds")
}
```

```{r, eval=FALSE}
cd4.best <- searchVarReg(y=CD4$cd4, 
                         x=CD4$age, 
                         maxknots.m = 7, 
                         maxknots.v = 7, 
                         selection="AIC", 
                         maxit=100) 
```


lets look at the AIC table to identify where the best model is located:

```{r}
cd4.best$AIC %>%
  kbl(digits=1)
```

The best model is saved within the `cd4.best` list. The key estimates from the best model are:


```{r}
cd4.best$best.model$knots.m
cd4.best$best.model$knots.v

cd4.best$best.model$mean
cd4.best$best.model$variance
```

ie. 6 knots in mean and 3 in variance, with parameter estimates as given. 
We can then plot this best model:

```{r, fig.dim=c(6,6)}
plotVarReg(cd4.best$best.model, 
           xlab = "Age in years", 
           ylab = "CD4 cell count")
```

From these residuals, there still appears to be deviations from
normality in the residuals, again seen in both the Q-Q plot and the
histogram. Therefore the incorporation of a shape parameter should be
investigated. 

### `lssVarReg` function

In order to fit models with a shape parameter, we utilise
the `lssVarReg()` function. Firstly, let us fit a constant shape parameter
with 6 knots in the mean and 3 knots in the variance, that is,


```{r, echo=FALSE}
#$Y_i \sim SN\left(\xi_0+\sum\limits_{p=1}^8 \xi_p s_{ip}, \omega_0+\sum\limits_{q=1}^6 \omega_q l_{iq}, \nu_0 \right), ~~~ i=1,2,...,609$
#where $s_{ip}$ are the B-spline basis functions ($p=1,2,...,8$) for six
#knots and $l_{ip}$ are the B-spline basis functions ($q=1,2,...,6$) for
#three knots, both based upon age in years.

if (file.exists("results_con.rds")) {
  con.shape <- readRDS("results_con.rds")


} else {
  con.shape <- lssVarReg(y=CD4$cd4, x=CD4$age, 
                locationmodel="semi", knots.l=6, 
                scale2model="semi", knots.sc=3,  
                shapemodel="constant",  
                para.space="positive",
                location.init = cd4.best$best.model$mean, 
                scale2.init = cd4.best$best.model$variance,
                int.maxit = 10000,
                maxit=1000 )  #time-consuming function
  saveRDS(con.shape, "results_con.rds")
}

if (file.exists("results_lin.rds")) {
  linear.shape <- readRDS("results_lin.rds")
} else {
  linear.shape <- lssVarReg(y=CD4$cd4, x=CD4$age, 
                locationmodel="semi", knots.l=6, 
                scale2model="semi", knots.sc=3,  
                shapemodel="linear",   
                para.space="positive",
                location.init = cd4.best$best.model$mean, 
                scale2.init = cd4.best$best.model$variance,
                maxit=1000 )  #time-consuming function
  saveRDS(linear.shape, "results_lin.rds")
}
```

Let us fit two models, one with a constant shape parameter and one with
a linear model in the shape, with the following code.

```{r, eval=FALSE}
con.shape<-lssVarReg(y=CD4$cd4, 
                     x=CD4$age, 
                     locationmodel="semi", 
                     knots.l=6, 
                     scale2model="semi", 
                     knots.sc=3, 
                     mono.scale = "inc", 
                     shapemodel="constant",   
                     maxit=1000 )

linear.shape<-lssVarReg(y=CD4$cd4, 
                        x=CD4$age, 
                        locationmodel="semi", 
                        knots.l=6, 
                        scale2model="semi", 
                        knots.sc=3,  
                        shapemodel="linear", 
                        int.maxit=1000, 
                        print.it=TRUE, 
                        maxit=1000 )
```

If we want to speed up the model, we can use starting estimates (from our best model) and parameter space restrictions, like so for the constant model:
```{r, eval=FALSE}
con.shape_faster <- lssVarReg(y=CD4$cd4, x=CD4$age, 
                              locationmodel="semi", 
                              knots.l=6, 
                              scale2model="semi", 
                              knots.sc=3,  
                              shapemodel="constant",  
                              para.space="positive", 
                              location.init = cd4.best$best.model$mean, 
                              scale2.init = cd4.best$best.model$variance,
                              int.maxit = 10000,
                              maxit=1000 )
```


And compare the models as we did in the paper:

```{r}
data.frame(Model = c("No shape", "Constant shape", "Linear shape"), 
           AIC = c(cd4.best$best.model$aic, con.shape$aic, linear.shape$aic), 
           BIC = c(cd4.best$best.model$bic, con.shape$bic, linear.shape$bic)) %>% 
  kbl(digits=1, 
      align='c', 
      caption = "Comparison of AIC from shape models for CD4 cell counts.") %>%
  kable_paper(full_width = TRUE ) 
```

And compare the residuals:

```{r}

n<-length(CD4$age)
knot6<-bs(CD4$age, df= 8, degree=2)
knot3<-bs(CD4$age, df= 5, degree=2)

##normal model
npred.mean<-colSums(t(cbind(rep(1,n),knot6))*cd4.best$best.model$mean)
npred.var<-colSums(t(cbind(rep(1,n),knot3))*cd4.best$best.model$variance)

stand.res<- (CD4$cd4-npred.mean)**2/(npred.var)
chiq<-qchisq(c(1:n)/(n+1), df=1)

#constant residuals
con.res<-lss_calc(con.shape)
con.res<-con.res[order(con.res$CD4.age),]

#linear residuals
lin.res<-lss_calc(linear.shape)
lin.res<-lin.res[order(lin.res$CD4.age),]

par(mfrow=c(1,3))
plot(chiq, sort(stand.res), main=NULL, ylim=c(0, 10), pch=19, ylab="Squared standardised residuals after fit with no shape", xlab="Chi-Square(1) quantiles")
abline(0,1)
mtext('A', side=3, line=2, at=0, adj=3)

plot(chiq, sort(con.res$stand.res2), main=NULL, ylim=c(0, 10),pch=19, ylab="Squared standardised residuals after LSS (constant) fit", xlab="Chi-Square(1) quantiles")
abline(0,1)
mtext('B', side=3, line=2, at=0, adj=3)
plot(chiq, sort(lin.res$stand.res2), main=NULL, ylim=c(0, 10),pch=19, ylab="Squared standardised residuals after LSS (linear) fit", xlab="Chi-Square(1) quantiles")
abline(0,1)
mtext('C', side=3, line=2, at=0, adj=3)
```


## Viral load dataset (censored outcome data)

This is a dataset of the HIV viral load (blood concentration of HIV RNA
on a log10 scale) measured in 257 participants. Prior to commencing a
clinical trial, participants had their blood assayed twice during a
short period of time. Although the underlying viral load is
unchanged in this time, the readings will differ due to measurement
error. Another important aspect is that measurements cannot be detected
below a particular assay limit, in this case, 2.70. 

Lets plot the data:
  
```{r}
ggplot(rna, aes(x=x, y=y))+
  geom_point()+
  geom_hline(yintercept = 0)+
  xlab("Mean viral load")+
  ylab("Difference in viral load")+
  theme_minimal()
```

Now let us search for the optimal knots, using the censoring indicator and 


```{r, echo=FALSE}
if (file.exists("RNAresults.rds")) {
  rna.best <- readRDS("RNAresults.rds")
} else {
  rna.best <- searchVarReg(y=rna$y, 
                           x=rna$x, 
                           cens.ind = rna$cc,
                           maxknots.m = 5,
                           maxknots.v = 5,
                           mono.var = "inc",  
                           selection="AIC", maxit=1000)  #time-consuming function
  saveRDS(rna.best, "RNAresults.rds")
}
if (file.exists("RNAresults2.rds")) {
  rna.best2 <- readRDS("RNAresults2.rds")
} else {
  rna.best2 <- searchVarReg(y=rna$y, x=rna$x, cens.ind = rna$cc,
                            maxknots.m = 5, maxknots.v = 5,
                            selection="AIC", maxit=1000)  #time-consuming function
  saveRDS(rna.best2, "RNAresults2.rds")
}
```

Monotonic decreasing model:
  
```{r, eval=FALSE}
rna.best <- searchVarReg(y=rna$y, 
                         x=rna$x, 
                         maxknots.m = 5, 
                         maxknots.v = 5,
                         mono.var = "dec",  
                         selection="AIC", 
                         maxit=1000)  
```

AIC from all models:

```{r}
rna.best$AIC %>%
  kbl(digits=1)
```


Find the smallest AIC:
```{r}
min(rna.best$AIC)

```

Allows increasing and decreasing:
  
```{r, eval=FALSE}
rna.best2 <- searchVarReg(y=rna$y, 
                          x=rna$x, 
                          cens.ind = rna$cc,
                          maxknots.m = 5, 
                          maxknots.v = 5,
                          selection="AIC", 
                          maxit=1000)
```

AIC from all models:

```{r}
rna.best2$AIC %>%
  kbl(digits=1)
```

Smallest AIC:
```{r}
min(rna.best2$AIC)
```

